NOPT048 Notes

### 2019-10-28 misc/study-notes/groups.tex
Studying for groups midterm

\def\Sym{\mathop{\rm Sym}\nolimits}
\def\Aut{\mathop{\rm Aut}\nolimits}
\def\Im{\mathop{\rm Im}\nolimits}
\def\Id{\mathop{\rm Id}\nolimits}

@@

If $G$ is a group containing an element $g$ such that $\langle g\rangle = G$, then $G$ is called a {\it cyclic group}.

@@

Formally, an {\it isomorphism} is a bijection $f:G\rightarrow G'$ from a group to another such that
$$f(x\cdot y) = f(x) \cdot f(y).$$
The multiplication on the left-hand side is taking place in $G$ while the right-hand multiplication takes place in $G'$. In the above example, the function $f : G_1 \rightarrow G_2$ that takes $i^k \mapsto \rho^k$ for $k=0,1,2,3$ gives an explicit isomorphism. If there exists an isomorphism between two groups, we say that they are {\it isomorphic}.

@@

An isomorphism from a set to itself is called an {\it automorphism}. Given a group $G$, we can construct a set $\Aut(G)$ of all automorphisms of $G$ and in fact, it is easily verifiable that $\Aut(G)$ is a group under function composition.

@@

Let $f:G\rightarrow G'$. The {\it kernel} of $f$, denoted $\ker(f)$, is a the set of all elements in the $G$ that map to the identity $e'$ of the target group $G'$. The {\it image} of $f$, denoted $\Im(f)$, is the set of all elements in $G'$ that equal $f(g)$ for some $g\in G$. Then the homomorphism $f$ is an isomorphism if and only if $\ker(f) = \{e\}$ and $\Im(f) = G'$. It is an easy exercise to verify that $\ker(f)$ and $\Im(f)$ are subgroups of $G$ and $G'$ respectively. In fact, the kernel of a homomorphism is a normal subgroup, as we shall now prove.

\proclaim Theorem K. Let $f:G\rightarrow G'$ be a group homomorphism. Then $K = \ker(f)$ is a normal subgroup of $G$.

\proof By Lemma 2.4.C, we need only show that $K$ is stable under conjugation. Let $k\in K$ be given and let us conjugate it by an element $a\in G$ to get $aka^{-1}$. Applying the homomorphism $f$ to $aka^{-1}$, we find that
$$f(aka^{-1}) = f(a)f(k)f(a^{-1}) = f(a)e'f(a)^{-1},$$
where $e'$ is the identity of $G'$. But $f(a)e'f(a)^{-1} = e'$, so $aka^{-1}\in \ker(f) = K$.\slug

@@

Let $G$ be a group and $S$ a set of elements. We say that $G$ {\it acts on} $S$ if there exists a map from $G\times S\rightarrow S$ (whose pairs $(g,s)$ are written $g(s)$) such that, for all $s\in S$, $e(s) = s$ and $(gh)(s) = g\big(h(s)\big)$ for every $g,h\in G$. For any element $s\in S$, the {\it orbit} is set of elements in $S$ to which $s$ may be taken by $G$:
$$O_s = \{g(s)\in S : g\in G\}$$
If there exists an $s$ such that $O_s = S$, then we say that $G$ acts {\it transitively} on $S$. In fact, because the orbits of a group action partition the set $S$ into equivalence classes, transitivity of a group action implies there is only one orbit, and for any $s,s'\in S$, there exists some $g\in G$ such that $g(s) = s'$.

>>>>

\def\Sym{\mathop{\rm Sym}\nolimits}
\def\Aut{\mathop{\rm Aut}\nolimits}
\def\Out{\mathop{\rm Out}\nolimits}
\def\Inn{\mathop{\rm Inn}\nolimits}
\def\Im{\mathop{\rm Im}\nolimits}
\def\Id{\mathop{\rm Id}\nolimits}

@@

If $G$ is a group containing an element $g$ such that $\langle g\rangle = G$, then $G$ is called {\it cyclic} or {\it singly-generated}. More generally, if $S\subset G$ and we let $S^{-1}$ denote the set $\{s^{-1} : s\in S\}$, then the set
$$\langle S\rangle = \{s_1s_2\cdots s_n : n\in\N, s_i\in S\cup S^{-1}\}$$
is a subgroup of $G$, called the subgroup generated by $S$. If it so happens that $\langle S\rangle = G$, then we way that $S$ is a {\it generating set} of $G$.

@@

Formally, an {\it isomorphism} is a bijection $f:G\rightarrow G'$ from a group to another such that
$$f(x\cdot y) = f(x) \cdot f(y).$$
The multiplication on the left-hand side is taking place in $G$ while the right-hand multiplication takes place in $G'$. In the above example, the function $f : G_1 \rightarrow G_2$ that takes $i^k \mapsto \rho^k$ for $k=0,1,2,3$ gives an explicit isomorphism. If there exists an isomorphism between two groups $G$ and $H$, we say that they are {\it isomorphic} and write $G\simeq H$.

@@

@@

\section 2.6. The First Isomorphism Theorem

Let $f:G\rightarrow H$ be a homomorphism. The {\it kernel} of $f$, denoted $\ker(f)$, is a the set of all elements in the $G$ that map to the identity $e_H$ of the target group $H$. The {\it image} of $f$, denoted $\Im(f)$, is the set of all elements in $H$ that equal $f(g)$ for some $g\in G$. In particular, the homomorphism $f$ is an isomorphism if and only if $\ker(f) = \{e\}$ and $\Im(f) = G$. It is an easy exercise to verify that $\ker(f)$ and $\Im(f)$ are always subgroups of $G$ and $H$ respectively.

Now we want to describe the preimages of $\Im(f)$. For each element $h\in Im(f)\leq H$, the subset $\{g : f(g) = h\}$ is its {\it fibre}. The following lemma establishes that the kernel of a homomorphism is normal and relates the fibres of the image to cosets of the kernel. Since the image of a homomorphism is a subgroup, we can assume that the homomorphism is surjective without losing generality.

\parenproclaim Lemma F (Fibre lemma). Let $f:G\rightarrow H$ be a surjective group homomorphism with kernel $K$. Then for any $h\in H$ and $g\in G$ that satisfies $f(g) = h$, we have $f^{-1}(h) = gK = Kg$, implying that $K$ is a normal subgroup of $G$.

\proof Let $h\in H$ and let $X$ denote $f^{-1}(h)$, the fibre of $h$. Now take $g\in G$ such that $f(g) = h$. Since $f(k) = e_H$ for all $k\in K$, we have $f(sk) = f(s)f(k) = h = f(k)f(s) = f(ks)$. So $sK\subset X$ and $Ks\subset H$. To prove the reverse inclusions, we let $x\in X$ (so $f(x) = h$). Then since $f$ is a homomorphism, we have $f(g^{-1}) = h^{-1}$ and
$$f(xg^{-1}) = f(x)f(g^{-1}) = hh^{-1} = e_H = h^{-1}h = f(g^{-1})f(x) = f(g^{-1}x).$$
This implies that $xg^{-1}\in K$, so multiplying on the right by $g$, we find that $x\in Kg$. Similarly, $g^{-1}x\in K$ so $x\in gK$.\slug

Now for any surjective map $f$ from sets $G$ to $H$, we can define an equivalence relation on elements $a,b\in G$:
$$a\sim b\quad \hbox{if and only if}\quad f(a) = f(b)$$
The set of equivalence classes is denoted $G/{\sim}$ and there exists a bijection $f':G/{\sim} \rightarrow H$. Now if $f$ is a group homomorphism, then we can pull back the group structure of $H$ to $G/{\sim}$ using the bijection. Then the set
$$G/K = \{gK : g\in G\}.$$
has a group structure and is called the {\it quotient group}. Multiplication in the quotient group is given by $gK\cdot hK = (g\cdot h) K$, which is well-defined because $K$ is normal. So the group structure of $G/K$ depends only on the kernel $K$ and not the homomorphism that induced it. The results of this section are summed up in the following theorem.

\parenproclaim Theorem I (First isomorphism theorem). Let $f:G\rightarrow H$ be a surjective group homomorphism. Then $K = \ker(f)$ is a normal subgroup of $G$ and there exists a group homomorphism $f':G/K \rightarrow H$.\slug

\section 2.7. Inner and Outer Automorphisms

An isomorphism from a group to itself is called an {\it automorphism}. Given a group $G$, we can construct a set $\Aut(G)$ of all automorphisms of $G$ and in fact, it is easily verifiable that $\Aut(G)$ is a group under function composition.

For any element $a$ of a group $G$, the map $f_a$ defined by
$$f_a(s) = asa^{-1}$$
is an automorphism. Such automorphisms, which conjugate the group by a fixed element, are called {\it inner} automorphisms. An automorphism that is not inner is called an {\it outer} automorphism. There is a homomorphism $\phi:G\rightarrow \Aut(G)$ given by $a\mapsto f_a$; that is, $f_af_b = f_{ab}$. This follows from associativity:
$$f_af_b(s) = a(bsb^{-1})a^{-1} = (ab)s(ab)^{-1} = f_{ab}(s)$$
The image of the homomorphism $\phi$ is denoted $\Inn(G)$.

The {\it centre} of a group $G$ is the subgroup
$$Z(G) = \{z\in G : zg = gz\ \hbox{for all}\ g\in G\}.$$

\proclaim Lemma Z. The centre of $G$ is the kernel of the map from $G$ to $\Aut(G)$ given by $a\mapsto f_a$.

\proof Let $a\in Z(G)$ and note that for an arbitrary element $s\in G$,
$$f_a(s) = asa^{-1} = aa^{-1}s = s$$
so $f_a$ is the identity automorphism. Likewise, if some $f_a\in \Aut(G)$ is the identity automorphism, then $f_a(s) = asa^{-1} = s$ for all $s\in G$. So $as = sa$ for all $s\in G$ and $a\in Z(G)$.\slug

From Lemma 2.4.C, we know that normal subgroups are exactly the subgroups that are stable under conjugacy, i.e.\ stable under inner automorphism. We can extend this into a more general definition. If $H\leq G$ and $f(H) = H$ for all $f\in \Aut(G)$, then we say that $H$ is a {\it characteristic} subgroup. (Of course, every characteristic subgroup is also normal.)

\proclaim Lemma C. The centre $Z(G)$ of a group $G$ is characteristic.

\proof It must be shown that if $g\in Z(G)$, then for all $f\in\Aut(G)$, $f(g)\in Z(G)$. We prove this by contraposition. Let $g\in G$ and suppose there exists $f\in\Aut(G)$ such that $f(g)\notin Z(G)$. Then there exists $s\in G$ such that $f(g)\cdot s \neq s\cdot f(g)$. Now we apply the inverse automorphism $f^{-1}$ to both sides to get that $g\cdot f^{-1}(s) \neq f^{-1}(s)\cdot g$. Hence there exists an element of $G$, namely $f^{-1}(s)$, that $g$ does not commute with, so $g\notin Z(G)$.\slug

By the first isomorphism theorem, $\Inn(G)\simeq G/Z(G)$; in particular, if $G$ is abelian, we have $Z(G) = G$ and $\Inn(G)$ is trivial. Because it arose as the image of a group homomorphism $\phi$, we know that $\Inn(G)\leq \Aut(G)$. In fact, we can prove that $\Inn(G)$ is normal in $\Aut(G)$.

\proclaim Lemma N. Let $G$ be a group. Then $\Inn(G)$ is a normal subgroup of $\Aut(G)$.

\proof Let $f_a\in Inn(G)$ be the map that conjugates by an element $a\in G$. Let $g\in \Aut(G)$ be arbitrary. Then
$$g\circ f_a\circ g^{-1} = g(a\cdot g^{-1}(s)\cdot a^{-1}) = g(a)\cdot s\cdot g(a)^{-1},$$
and we see that $g\circ f_a \circ g^{-1} = f_{g(a)}\in Inn(G)$. So $\Inn(G)$ is stable under conjugation in $Aut(G)$.\slug

Now by the first isomorphism theorem, we have a group $\Out(G) = \Aut(G)/\Inn(G)$, called the {\it outer automorphism group}. Note that this is not the set of outer automorphisms! (The set of outer automorphisms do not form a group because the identity automorphism is inner.)

\section ?.?. Simplicity of the Alternating Group

A group $G$ is {\it simple} if its only normal subgroups are the trivial group $\{e\}$ and the whole group $G$. In this section we show that for $n\geq 5$, the alternating group $A_n$ is simple. First we establish that the set of 3-cycles generate $A_n$.

\proclaim Lemma G. Every permutation in $A_n$ can be written as a product of 3-cycles.

\proof Since every permutation in $A_n$ can be written as a product of an even number of transpositions, it suffices to show that any product $(a\,b)(c\,d)$ can be rewritten as a product of 3-cycles. There are three cases to consider. Firstly, if the transpositions have two numbers in common, then $(a\,b) = (c\,d)$ and the product is simply the identity permutation. If the transpositions have a single point in common, then the product is a 3-cycle, since $(a\,b)(b\,c) = (a\,b\,c)$. Lastly, in the case that all four points are distinct, we have
$$(a\,b)(c\,d) = (a\,b\,c)(b\,c\,d),$$
a product of 3-cycles.
Hence the set of 3-cycles generates $A_n$.\slug

In fact, we can shrink the generating set to 3-cycles of a specific form.

\proclaim Lemma H. The $n-2$ permutations $(1\,2\,3), (1\,2\,4),\ldots,(1\,2\,n)$ generate $A_n$.

\proof By Lemma G, we need only show that any 3-cycle can be rewritten as a product of cycles $(1\,2\,k)$. Since the square of $(1\,2\,k)$ is $(2\,1\,k)$, any cycle that contains both 1 and 2 can be generated by the permutations above. Then, note that any cycle $(a\,b\,c)$ that fixes 1 can be rewritten $(1\,a\,b)(1\,b\,c)$. Now we can rewrite each of these cycles as follows:
$$(1\,j\,k) = (1\,2\,k)(1\,2\,k)(1\,2\,j)(1\,2\,k)$$
So every 3-cycle can be expressed as a product of cycles of the form $(1\,2\,k)$.\slug

This result can be used to show that any normal subgroup of $A_n$ that contains a 3-cycles is necessarily the whole group $A_n$.

\proclaim Lemma K. Let $n\geq 4$ and let $N$ be a normal subgroup of the group $A_n$. If $N$ contains a 3-cycle, then $N = A_n$.

\proof Without loss of generality, suppose that $N$ contains the cycle $(1\,2\,3)$. Then it also contains its inverse $(2\,1\,3)$ and for any $\sigma\in A_n$,
$$\sigma (2\,1\,3) \sigma^{-1}\in N.$$
So let $\sigma = (1\,2)(3\,k)$ for $k\geq 4$ and we have
$$\sigma (2\,1\,3) \sigma^{-1} = (1\,2\,k).$$
By Lemma H, these generate $A_n$, so $N$ is the entire group.\slug

We may finally prove that $A_n$ is simple for $n\geq 5$. The following theorem is due to E. Galois.

\proclaim Theorem S. Let $n\geq 5$ and let $N\geq A_n$ be non-trivial. Then $N=A_n$.

\proof Let $\tau\in N$ be a non-identity permutation that fixes as many symbols as possible. We will show, by contradiction, that $\tau$ must be a 3-cycle, i.e.\ it fixes all symbols except three.

Suppose that $\tau$ moves more than three symbols. Then either $\tau$ has a cycle with more than three symbols, or $\tau$ consists of disjoint transpositions. Without losing generality, we use the symbols 1, 2, 3, 4, and 5. In the first case, we have
$$\tau = (1\,2\,3\,\ldots)\ldots$$
and we can conjugate by $(3\,4\,5)$ to get
$$\tau' = (3\,4\,5)\tau(3\,4\,5)^{-1} = (1\,2\,4\,\ldots)\ldots.$$
Then $\tau^{-1}\tau'$ fixes 1 where $\tau$ did not, a contradiction.

In the case that $\tau$ only contains transpositions, we have
$$\tau = (1\,2)(3\,4)\ldots,$$
and again we conjugate by $(3\,4\,5)$ to get
$$\tau' = (3\,4\,5)\tau(3\,4\,5)^{-1} = (1\,2)(4\,5)\ldots)\ldots.$$
Then $\tau^{-1}\tau'$ fixes 1 and 2 and $\tau$ does not, leading to a contradiction.

The contradiction in both cases implies that $\tau$ moves three symbols or less. In the alternating group, this implies that $\tau$ is a 3-cycle. So $N$ contains a 3-cycle and by Lemma K, $N=A_n$.\slug

@@

Let $G$ be a group and $S$ a set of elements. We say that $G$ {\it acts on} $S$ if there exists a map from $G\times S\rightarrow S$ (whose pairs $(g,s)$ are written $g(s)$ or sometimes $gs$) such that, for all $s\in S$, $e(s) = s$ and $(gh)(s) = g\big(h(s)\big)$ for every $g,h\in G$. For any element $s\in S$, the {\it orbit} is set of elements in $S$ to which $s$ may be taken by $G$:
$$O_s = \{g(s)\in S : g\in G\}$$
If there exists an $s$ such that $O_s = S$, then we say that $G$ acts {\it transitively} on $S$. In fact, because the orbits of a group action partition the set $S$ into equivalence classes, transitivity of a group action implies there is only one orbit, and for any $s,s'\in S$, there exists some $g\in G$ such that $g(s) = s'$.
